Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model

### 2023-05-16 src/docs_paper/appendix.Rmd
Check in paper and appendix writing

We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
Let $u$ be a generic random effect, with length $\dim(u)$.
Three types of structured random effects are used in the model.
First, we specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(u)
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the lag-one correlation parameter.
Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th element removed i.e. $(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
Third, for the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v + \sqrt{1 - \phi} \cdot w \right).
$$

@@

We assume the household survey was run according to a complex survey design where each individual $j \in U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then the weighted mean
$$
\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
$$
We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
$$
m^{\hat \theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
$$
The observed number of indicator cases is then $y^{\hat \theta} = m^{\hat \theta} \cdot \hat \theta$.
To make these computations we use the `survey` \textsc{R} package [@lumley2004analysis].

>>>>

We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as those belonging to neighbouring districts or adjacent age-bands.
Let $u$ be a generic random effect, with length $\dim(u)$.
Three types of structured random effects are used in the model.
First, we specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_i, \quad i = 2, \ldots, \dim(u)
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ are independent and identically distributed (IID) Gaussian random variables, $\sigma > 0$ is the marginal standard deviation, and $|\rho| < 1$ is the (lag-one) correlation parameter.
Second, we use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th element removed i.e. $(u_1, \ldots, u_{i - 1}, u_{i + 1}, \ldots, u_{\dim(u)})$, $j \sim i$ if the units $i$ and $j$ are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
Third, for the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising], we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v + \sqrt{1 - \phi} \cdot w \right).
$$

@@

We assume the household survey was run according to a complex survey design where each individual $j$ in the population $U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $\theta_j \in \{0, 1\}$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then a survey weighted mean is given by
$$
\hat \theta = \frac{\sum_{j \in S} w_j \theta_j}{\sum_{j \in S} w_j}.
$$
We account for survey weighting in the variance via the Kish effective sample size [@kish1965survey]
$$
m^{\hat \theta} = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
$$
The observed number of indicator cases is then $y^{\hat \theta} = m^{\hat \theta} \cdot \hat \theta$.
To make these computations we use the `survey` \textsc{R} package [@lumley2004analysis].

### 2023-05-16 src/docs_paper/paper.Rmd
Check in paper and appendix writing

@eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points $T_i, i = 1, 2, 3$.
We consider a simplified version defined only at the time of the most recent household survey with HIV testing ($T_1$), omitting nowcasting ($T_2$) and temporal projection ($T_3$), as these time points involve limited additional inference.
An overview of this simplified model is given below.
A more complete mathematical description (Appendix S1) as well as a \textsf{C++} template for the log-posterior (Appendix S3) are provided in the supplementary material.

@@

Let $N_i \in \mathbb{N}$ be the known, fixed population size.
We infer the following unknown HIV indicators: HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that
\begin{align*}
\text{logit}(\rho_i) &= \eta^\rho_i, \\
\text{logit}(\alpha_i) &= \eta^\alpha_i,
\end{align*}
for certain choice of linear predictors $\eta^\rho_i$ and $\eta^\alpha_i$.
HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i(\{\rho_i, \alpha_i\}_{i \in \mathcal{I}})$, where the linear predictor depends on $\{\rho_i, \alpha_i\}_{i \in \mathcal{I}}$ for some $\mathcal{I}$.
Finally, let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
$$
\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right),
$$
where the mean duration of recent infection $\Omega_T$ and the proportion of long-term HIV infections misclassified as recent $\beta_T$ are strongly informed by priors for the particular survey.

For $\theta \in \{\rho, \alpha, \kappa\}$ we calculate the weighted, aggregated survey observations
$$
\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j},
$$
where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
We calculate the observed number of outcomes as $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
$$
m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
is the Kish effective sample size [@kish1965survey].
We use a binomial working likelihood
$$
y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})
$$
to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following weighted aggregates
\begin{equation*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \quad
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \quad
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{equation*}

@@

HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset on the logit scale from the general population indicators as follows
\begin{align*}
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\alpha^\text{ANC}}_i.
\end{align*}
These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
In particular, the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
We use the binomial working likelihoods
\begin{align*}
y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(x^\text{ANC}_\mathcal{I}, \rho^\text{ANC}_{\mathcal{I}}), \\
z^\text{ANC}_\mathcal{I} &\sim \text{Bin}(y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}),
\end{align*}
where, again, we use weighted aggregates
\begin{equation*}
\rho^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in \mathcal{I}} \Psi_i}, \quad
\alpha^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}},
\end{equation*}
with $\Psi_i$ the number of pregnant women, which we assume to be fixed.

<!-- ## ART attendance component \label{sec:art} -->

<!-- We use a multinomial logistic regression model to account for people living with HIV choosing to access ART services outside of the district that they reside in. -->
<!-- Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$. -->
<!-- We assume that $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are adjacent such that $x \sim x'$.  -->
<!-- We model the log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ using a linear predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely. -->
<!-- We model aggregate ART attendance data $\dot A_\mathcal{I}$ using a Gaussian approximation to a sum of binomials. -->
<!-- The sum results both by aggregation over $i \in \mathcal{I}$ and by number of ART clients travelling from district $x'$ to $x$. -->
<!-- More details regarding this part of the model are provided in Appendix S1. -->

@@

@@

Extended latent Gaussian models (ELGMs) [@stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this mapping allows for the presence of more non-linearity in the model.
ELGMs are then of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n].
\end{align*}

## Naomi

The Naomi model is not an LGM, and instead falls into the ELGM class.

<!-- 1. In the ANC testing component (Section \ref{sec:anc}), the HIV prevalence and ART coverage depend upon the household survey component. -->
<!-- Specifically, $|\mathcal{J}_i| = 2$ such that for $\theta \in \{\rho, \alpha\}$ -->
<!-- $$ -->
<!-- \mu_i = g_i(\eta^\theta_i, \eta^{\theta^{\text{ANC}}}_i) = \text{logit}^{-1}(\eta^\theta_i + \eta^{\theta^{\text{ANC}}}_i). -->
<!-- $$ -->
<!-- 2. In the household survey component (Section \ref{sec:household}), the HIV incidence rate depends on adult HIV prevalence and adult ART coverage via non-linear transformations. -->
<!-- 3. Individuals who are taking ART, or have been recently infected, must be HIV positive. As a result observations are made taking the product of HIV prevalence with ART coverge or recent HIV infection. -->
<!-- 4. Recent HIV infection is non-linearly linked to both HIV incidence and HIV prevalence. -->
<!-- 5. The ART attendance component (Section \ref{sec:art}) uses a multinomial model. -->
<!-- 6. Processes are modelled at the finest distict-age-sex division $i$, but likelihoods are defined for observations aggregated over sets of indices $\mathcal{I}$. -->
<!-- 7. Multiple link functions are used: there is no one $g$. -->

@@

To write.

@@

A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by the following weighted sum
$$
\int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z),
$$
where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials, and weights so as to interpolate functions of the form $f(\z) = \tilde f(\z) \exp(-\z^2)$.
These nodes and weights can be shifted and scaled to suit the particular integrand using adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate].
AGHQ with $k = 1$ recovers the Laplace approximation.

```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension are adapted based on the mean and covariance matrix of the target via $\\btheta(\\z) = \\hat \\btheta + \\mathbf{L} \\z$."}
mu <- c(1, 1.5)
cov <- matrix(c(2, 1, 1, 1), ncol = 2)

@@

plot1 <- add_points(plot0, gg) +
  labs(size = "", caption = "")

@@

plot2 <- add_points(plot0, gg2) +
  labs(size = "", caption = "")

@@

plot3 <- add_points(plot0, gg3) +
  labs(size = "", caption = "")

plot1 + plot2 + plot3
```

## Algorithm\label{sec:algorithm}

@@

We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using four inferential approaches.
For each approach, the `TMB` C++ user-template (available in the appendix) used to specify the log-posterior was the same.
The four approaches were:
1. `TMB`: EB combined with a Gaussian approximation via `TMB`,
2. `aghq`: AGHQ combined with a Gaussian approximation via `aghq`,
3. `adam`: AGHQ combined with a Laplace approximation by extending `aghq`, and
4. `NUTS`: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
In this instance, the dimension of the latent field is $N = 491$ and the dimension of the hyperparameters is $m = 24$.
Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
Example model outputs from `TMB` are illustrated in Figure \ref{fig:naomi-results}.
The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].

@@

\begin{table}[]
\small
\begin{tabularx}{\textwidth}{p{0.3\linewidth}p{0.65\linewidth}}
\toprule
Inferential method & Details \\
\midrule
1. \texttt{TMB}: EB, Gaussian & $1000$ samples \\
2. \texttt{aghq}: AGHQ, Gaussian & $k = 1$, $1000$ samples \\
3. \texttt{adam}: AGHQ, Laplace & $k = 1$, $l = 5$, $1000$ samples \\
4. \texttt{NUTS}: NUTS & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$. HMC parameters set to default for \texttt{rstan}. \\
\bottomrule
\end{tabularx}
\caption{A summary of settings used for each inferential method.}
\label{tab:inference-methods}
\small
\end{table}

@@

Inferential results from MCMC are only accurate once convergence has been reached and the chain lengths are sufficient.
We assessed the quality of our MCMC results using the potential scale reduction factor $\hat R$, bulk and tail effective sample size (ESS), autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
Full details are provided in the appendix.
We treat these results from NUTS as a gold-standard to which other inferential methods are compared.

>>>>

@eaton2021naomi specify a joint model linking three small-area estimation models, defined over three time points $T_i, i = 1, 2, 3$.
We consider a simplified version defined only at the time of the most recent household survey with HIV testing ($T_1$), omitting nowcasting ($T_2$) and temporal projection ($T_3$) which involve limited additional inference.
An overview of this simplified model is given below.
A more complete mathematical description (Appendix S1) as well as a \textsf{C++} template for the log-posterior (Appendix S3) are provided in the supplementary material.

@@

Let $N_i \in \mathbb{N}$ be the known, fixed population size.
We infer the following unknown HIV indicators using coupled regression equations: HIV prevalence $\rho_i \in [0, 1]$, the proportion of individuals who are HIV positive; antiretroviral therapy (ART) coverage $\alpha_i \in [0, 1]$, the proportion of people living with HIV who receive ART treatment; and annual HIV incidence rate $\lambda_i > 0$, the yearly rate of new HIV infections occurring.
Independent logistic regression models for HIV prevalence and ART coverage in the general population are specified such that $\text{logit}(\rho_i) = \eta^\rho_i$ and $\text{logit}(\alpha_i) = \eta^\alpha_i$, for certain choice of structured additive predictors.
HIV incidence rate is modelled on the log scale as $\log(\lambda_i) = \eta^\lambda_i$, and depends on adult HIV prevalence and adult ART coverage. 
Let $\kappa_i$ be the proportion recently infected among HIV positive persons, which we link to HIV incidence via
\begin{equation}
\kappa_i = 1- \exp \left( - \lambda_i \cdot \frac{1 - \rho_i}{\rho_i} \cdot (\Omega_T - \beta_T) - \beta_T \right), \label{eq:kappa}
\end{equation}
where the mean duration of recent infection $\Omega_T$ and the proportion of long-term HIV infections misclassified as recent $\beta_T$ are strongly informed by priors for the particular survey.

These processes are informed by household survey data.
For $\theta \in \{\rho, \alpha, \kappa\}$ let
$$
\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}
$$
be weighted, aggregate survey observations, where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
The observed number of outcomes are $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
$$
m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
is the Kish effective sample size [@kish1965survey].
We use a binomial working likelihood
$$
y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})
$$
to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following weighted aggregates
\begin{equation*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \quad
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \quad
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{equation*}

@@

HIV prevalence $\rho^\text{ANC}_i$ and ART coverage $\alpha^\text{ANC}_i$ among pregnant women are modelled as offset from the general population indicators as follows
\begin{align*}
\text{logit}(\rho^\text{ANC}_i) &= \text{logit}(\rho_i) + \eta^{\rho^\text{ANC}}_i, \\
\text{logit}(\alpha^\text{ANC}_i) &= \text{logit}(\alpha_i) + \eta^{\alpha^\text{ANC}}_i.
\end{align*}
These processes are informed by likelihoods specified for aggregate ANC data from the year of the most recent survey.
In particular, the number of ANC clients with ascertained status $x^\text{ANC}_\mathcal{I}$, the number of those with positive status $y^\text{ANC}_\mathcal{I}$, and the number of ANC clients already on ART prior to their first ANC visit $z^\text{ANC}_\mathcal{I}$.
We use the binomial working likelihoods
\begin{align*}
y^\text{ANC}_\mathcal{I} &\sim \text{Bin}(x^\text{ANC}_\mathcal{I}, \rho^\text{ANC}_{\mathcal{I}}), \\
z^\text{ANC}_\mathcal{I} &\sim \text{Bin}(y^\text{ANC}_\mathcal{I}, \alpha^\text{ANC}_{\mathcal{I}}),
\end{align*}
where, again, we use weighted aggregates
\begin{equation*}
\rho^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}}{\sum_{i \in \mathcal{I}} \Psi_i}, \quad
\alpha^\text{ANC}_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC} \alpha^\text{ANC}_i}{\sum_{i \in \mathcal{I}} \Psi_i \rho_i^\text{ANC}},
\end{equation*}
with $\Psi_i$ the number of pregnant women, which we assume to be fixed.

## ART attendance component \label{sec:art}

People living with HIV sometimes choose to access ART services outside of the district that they reside in.
To account for this, we use multinomial logistic regression to model the probabilities of accessing services outside the home district.
Let $\gamma_{x, x'}$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$, and assume $\gamma_{x, x'} = 0$ unless $x = x'$ or the two districts are neighbouring, denoted by $x \sim x'$.
The log-odds $\tilde \gamma_{x, x'} = \text{logit}(\gamma_{x, x'})$ are modelled using a structured additive predictor $\eta_x^{\tilde \gamma}$ which only depends on the home district $x$, such that travel to each neighbouring district, for all age-sex strata, is equally likely.
Aggregate ART attendance data $\dot A_\mathcal{I}$ is modelled using a Gaussian approximation to a sum of binomials.
The sum results both by aggregation over $i \in \mathcal{I}$ and by number of ART clients travelling from district $x'$ to $x$.
More details regarding this part of the model are provided in Appendix S1.

@@

We now describe a popular class of models, and an extension encapsulates the complexities of Naomi.

@@

Extended latent Gaussian models (ELGMs) [@stringer2022fast] relax the restriction that there is a one-to-one mapping between the mean response $\bmu$ and structured additive predictor $\bmeta$.
Instead, the structured additive predictor is redefined as $\bmeta = (\eta)_{i \in [N_n]}$, where $N_n \in \mathbb{N}$ is a function of $n$, and it is possible that $N_n \neq n$.
Each mean response $\mu_i$ now depends on some subset $\mathcal{J}_i \subseteq [N_n]$ of indices of $\bmeta$, with $\cup_{i = 1}^n \mathcal{J}_i = [N_n]$ and $1 \leq |\mathcal{J}_i| \leq N_n$.
The inverse link function $g(\cdot)$ is redefined for each observation to be a possibly many-to-one mapping $g_i: \mathbb{R}^{|\mathcal{J}_i|} \to \mathbb{R}$, such that $\mu_i = g_i(\bmeta_{\mathcal{J}_i})$.
Importantly, this mapping allows for the presence of non-linearity in the model.
Put together, ELGMs are then of the form
\begin{align*}
y_i &\sim p(y_i \, | \, \bmeta_{\mathcal{J}_i}, \btheta_1), \quad i \in [n] \\
\mu_i &= \mathbb{E}(y_i \, | \, \bmeta_{\mathcal{J}_i}) = g_i(\bmeta_{\mathcal{J}_i}), \\
\eta_j &= \beta_0 + \sum_{l = 1}^{p} \beta_j z_{ji} + \sum_{k = 1}^{r} f_k(u_{ki}), \quad j \in [N_n].
\end{align*}

## Naomi as an ELGM

Naomi has a lot in common with many LGMS: it is a spatio-temporal model with a large latent field, governed by a smaller number of hyperparameters.
However, Naomi is not an LGM, and instead falls into the ELGM class, for the following reasons:

1. In the household survey component, HIV incidence depends on district-level adult HIV prevalence and ART coverage, such that $\lambda \propto \rho (1 - \omega \cdot \alpha)$, where $\omega = 0.7$ is a fixed constant. This reflects basic HIV epidemiology: HIV incidence is proportional to unsuppresed viral load. As a result, $\log(\lambda_i)$ depends on 28 structured additive predictors (2 sexes $\times$ 7 age groups $\times$ 2 indicators, HIV prevalence and ART coverage).
2. In the household survey component, HIV incidence and HIV prevalence are linked to the proportion recently infected via Equation \ref{eq:kappa}.
3. In the ANC testing component, HIV prevalence and ART coverage depend upon the respective indicators in the household survey component. Although $\text{logit}(\rho_i)$ and $\text{logit}(\alpha_i)$ are Gaussian, nonetheless this introduces dependence of each mean response on two structured additive predictors.
4. Throughout the model components, processes are modelled at the finest distict-age-sex division, but likelihoods are defined for observations aggregated over sets of indices. As such, single observations are related to $|\mathcal{I}|$ structured additive predictors.
5. Individuals taking ART, or who have been recently infected, must be HIV positive. 
6. The ART attendance component uses a multinomial model with softmax link function which takes as input $|\{x': x' \sim x\}| + 1$ structured additive predictors.
7. Multiple link functions are used throughout the model, such that there is no one inverse link function $g$. Instead,

@@

Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
Then the Laplace approximation to $p(\btheta, \y)$ is given by
$$
\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$

@@

A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by the following weighted sum
$$
\int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z),
$$
where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials, with weights chosen so as to interpolate functions of the form $f(\z) = \tilde f(\z) \exp(-\z^2)$.
In adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate] the nodes and weights are shifted and rotated to suit the particular integrand, based on the mode, curvature at the mode, and a matrix decomposition of the inverse curvature.
Two possibilities for this matrix decomposition are the Cholesky decomposition and the spectral decomposition [@jackel2005note].

```{r aghq, fig.cap="The Gauss-Hermite quadrature nodes $\\z \\in \\mathcal{Q}(2, 3)$ for this two dimensional integral with three nodes per dimension (a) are adapted based on the mean (b) and covariance matrix of the target via $f(\\z) = \\hat f + \\mathbf{L} \\z$, where..."}
mu <- c(1, 1.5)
cov <- matrix(c(2, 1, 1, 1), ncol = 2)

@@

plot1 <- add_points(plot0, gg) +
  labs(size = "", subtitle = "(a)")

@@

plot2 <- add_points(plot0, gg2) +
  labs(size = "", subtitle = "(b)")

@@

plot3 <- add_points(plot0, gg3) +
  labs(size = "", subtitle = "(c)")

#' Adapt by the spectral
gg4 <- gg
mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1)

plot4 <- add_points(plot0, gg4) +
  labs(size = "", subtitle = "(d)")

(plot1 + plot2) / (plot3 + plot4)
```

@@

We fit the simplified Naomi model (Section \ref{sec:naomi}) to data from Malawi using three inferential approaches.
For each approach, the `TMB` C++ user-template (available in the appendix) used to specify the log-posterior was the same.
The four approaches were:
1. TMB: EB combined with a Gaussian approximation via `TMB`,
2. PCA-AGQH: the PCA AGHQ grid combined with a Gaussian approximation via `aghq`,
3. NUTS: the Hamiltonian Monte Carlo (HMC) algorithm No-U-Turn Sampling (NUTS) using Stan [@carpenter2017stan] via the `tmbstan` package [@monnahan2018no].
The dimension of the latent field is $N = 491$ and the dimension of the hyperparameters is $m = 24$.
Settings used for each inferential method are provided in Table \ref{tab:inference-methods}.
For the deterministic methods, following inference we simulated hyperparameter and latent field samples.
For all methods, we simulated age-sex-district specific HIV prevalence, ART coverage and HIV incidence from the latent field and hyperparameter posteriors.
<!-- Note: eventually this figure should be generated using the algorithm proposed in the paper -->
Example model outputs from `TMB` are illustrated in Figure \ref{fig:naomi-results}.
The \textsc{R} [@r] code used to produce all results we describe below is available at `github.com/athowes/elgm-inf`.
We used `orderly` [@orderly] for reproducible research, `ggplot2` for data visualisation [@wickham2016ggplot2] and `rticles` [@allaire2022rticles] for reporting via `rmarkdown` [@allaire2022rmarkdown].

@@

\begin{table}[]
\small
\begin{tabularx}{\textwidth}{p{0.15\linewidth}p{0.15\linewidth}p{0.6\linewidth}}
\toprule
Name & Software & Details \\
\midrule
TMB & \texttt{TMB} & Empirical Bayes, $1000$ samples \\
PCA-AGHQ & \texttt{aghq} & $k = 3, s = 8$, $1000$ samples \\
NUTS & \texttt{tmbstan} & $4$ chains of $20000$ iterations with the first $10000$ iterations of each chain discarded as warmup, then thinned by a factor of $20$. HMC parameters set to default for software. \\
\bottomrule
\end{tabularx}
\caption{A summary of settings used for each inferential method.}
\label{tab:inference-methods}
\small
\end{table}

@@

We assessed the quality of our NUTS results using the potential scale reduction factor $\hat R$, bulk and tail effective sample sizes, autocorrelation decay plots, univariate traceplots, pairs density plots, and NUTS specific divergent transition and energy assessments.
Full details are provided in the appendix.
We  treat these results from NUTS as a gold-standard to which other inferential methods are compared.
